---
title: \vspace{-1.5cm} Causal Inference - Assignment Part1
author: 
- "Emilia Löscher" 
- "Kyuri Park"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
#bibliography: references.bib
---
\fontsize{11}{15}
\selectfont

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(qgraph)
library(dagitty)
library(ggdag)
library(dplyr)
library(pcalg)
```

## 1. Draw a DAG representing a simple and fairly plausible causal system from your preferred topic of choice. Describe briefly the substantive motivation behind your DAG. 
Many people with Obsessive-compulsive disorder (OCD) become depressed. According to Millet et al. (2004), the lifetime rates of major depression in OCD patients is about 81.2%. Several studies attempted to identify the causal mechanism in the comorbidity of OCD and depression (McNally et al.2017; Zandberg et al. 2015) and speculated that OCD symptoms often precede and correspondingly activate the depression symptoms.  

Given this background, here we proceed to look into the causal relationship between OCD and depression symptoms. We hypothesize that *distress associated with obsession* (OCD symptom) causes *feeling of guilt* (depression symptom) via several paths. Having been inspired by the DAG model from McNally et al. (2017), we construct our DAG with total 7 variables (nodes), which consist of OCD symptoms as well as depression symptoms. The specifics of the variables in our DAG is as follows:  

< OCD symptoms >

 - **ocdis**: distress caused by obsessions/compulsions
 - **ocint**: interference due to obsessions/compulsions
 - **occon**: difficulty controlling obsessions/compulsions

< Depression symptoms >  

 - **sad**: sadness 
 - **insom**: insomnia/sleeping problems 
 - **concen**: concentration/decision-making impairment
 - **guilt**: guilt and self-blame 

<!-- We chose the topic of obsessive-compulsive disorder (OCD) and it's relationship to depression symtoms. Our DAG is inspired by the paper "Co-morbid obsessive–compulsive disorder and depression: a Bayesian network approach" by R. J. McNally, P. Mair, B. L. Mugno and B. C. Riemann published in 2017. We broke down the DAG presented on page 1211 to only include 7 variables and constructed the variables according to the DAG and common sense.   -->

```{r plotdag, out.width="90%", fig.cap="OCD-Depression DAG", fig.align='center'}

# varnames <- c("ocint", "ocdis", "occon", "sad", "insom", "concen", "guilt")
# Adj <- matrix(c(0,1,0,0,0,1,0,
#                 0,0,1,1,0,0,0,
#                 0,0,0,1,0,0,0,
#                 0,0,0,0,1,0,1,
#                 0,0,0,0,0,1,0,
#                 0,0,0,0,0,0,1,
#                 0,0,0,0,0,0,0), 7,7, byrow = TRUE, dimnames = list(varnames, varnames))
# 
# qgraph(Adj, labels = varnames, layout= "groups", theme="classic")

ocddep <- dagify(
  ocdis ~ ocint,
  occon ~ ocdis,
  sad ~ ocdis + occon,
  insom ~ sad,
  concen ~ ocint + insom,
  guilt ~ concen + sad,
  exposure = "ocdis", # cause variable we are interested in
  outcome = "guilt", # effect variable we are interested in
  # set the coordinates
  coords = list(x = c(ocdis = -1, ocint=-0.5, occon=-0.5, sad = 0, insom=0, concen=0.5, guilt = 1),
              y = c(ocdis = 0, ocint=1, occon=-1, sad = 0, insom=0.5, concen=1, guilt = 0)) 
)

ggdag_status(ocddep) + theme_dag()

```

## 2. Specify a structural causal model for your DAG. Assume all relationships between the variables are linear, and all the variables have nrmally distributed residuals.  

In the following, the structural causal model for our DAG is specified:


$$ocint  := \epsilon_{ocint}$$
$$ocdis := 3 \cdot ocint + \epsilon_{ocdis}$$
$$occon := 4 \cdot ocdis + \epsilon_{occon}$$
$$sad := 2 \cdot ocdis + 1 \cdot occon + \epsilon_{sad}$$
$$insom := 2 \cdot sad + \epsilon_{insom}$$ 
$$concen :=  1 \cdot ocint + 4 \cdot insom + \epsilon_{concen}$$
$$guilt := 1 \cdot concen + 1 \cdot sad + \epsilon_{guilt}$$
,where $\epsilon_{ocint}, \ldots, \epsilon_{guilt} \stackrel{iid}{\sim} N(0 , SD_{i})$ with $i \in {ocint, \ldots, guilt}$.  

We base our estimates of the intercepts and residual standard deviations (SD) on the data from McNally et al. (2017), where they interviewed 408 patients for the OCD and depression symptoms. Intercepts and residual SDs are set to the mean and SD values of the corresponding item (symptom) that they found. The regression coefficients are decided such that it could represent the relative strength (effect) of each item we have thought of, considering the size of SD for each item.


## 3. Generate data from your SCM with a sample size of 500 units.  

```{r}
# set the seed
set.seed(1000)

# sample size (n) = 500
n <- 500

# generate the data
ocint <- 2.69 + rnorm(n, 0, 0.82)
ocdis <- 2.81 + 3.5 * ocint + rnorm(n, 0, 0.76)
occon <- 2.67 + 3 * ocdis + rnorm(n, 0, 0.76)
sad <- 1.55 + 4.3 * ocdis + 2 * occon + rnorm(n, 0, 0.94)
insom <- 0.81 + 2 * sad + rnorm(n, 0, 1.07)
concen <-  1.48 + 2 * ocint + 3 * insom + rnorm(n, 0, 0.87)
guilt <- 1.56 + 1.5 * concen + 3.3 * sad + rnorm(n, 0, 1.17)

# putting them in a dataframe called "OCDDEP"
OCDDEP <- data.frame(ocint, ocdis, occon, sad, insom, concen, guilt)
```

## 4. Use the PC-algorithm on the generated data to ‘discover’ the structure of your causal system

### a. Is your true DAG covered in the Markov equivalence class provided by the algorithm?
No, our true DAG is not covered in the Markov equivalence class. A couple of edges are missing (e.g., *ocint* --> *concen*, *ocdis* --> *occon*).

```{r pcalg, message=FALSE, out.width="90%", fig.cap="Inferred CPDAG using pcalg", fig.align='center'}
# sufficient statistics: correlation matrix + sample size
suffStat <- list(C= cor(OCDDEP), n = nrow(OCDDEP))

# run the pc algorithm
varnames <- colnames(OCDDEP)
pc_fit <- pc(suffStat = suffStat, indepTest = gaussCItest, alpha = 0.01, 
              labels = varnames)

# # plot the Markov equivalence class 
# if (require(Rgraphviz)) { 
#   p1 <- plot(pc_fit, main = "Inferred CPDAG using PC algorithm")
# }
laymat <- rbind(c(-0.5, 1),
                c(-1, 0),
                c(-0.5, -1),
                c(0, 0),
                c(0, 0.5),
                c(0.5, 1),
                c(1, 0))

par(mar=c(5, 4, 3.5, 2) + 0.1)

cpdag_mat <- as(pc_fit, "matrix")
qgraph(t(cpdag_mat), bidirectional=TRUE, color=c("white","salmon", "white", 
                                                 "white", "white", "white", "lightblue"), 
       layout = laymat, labels = varnames)
title("Inferred CPDAG using PC algorithm", line=3, cex.main = 1) 
```



### b. Provide the CP-DAG. To what extent did the procedure correctly recover which relationship were absent/present/directed?

```{r markov-equivalence set, fig.cap="Markov-Equivalence Set", fig.align='center'}
# extract the DAG adjacency matrix in a vector form (by rows)
res1 <- pdag2allDags(cpdag_mat)

# get the adjacency matrix of an individual DAG 
res1_dags <- list()
for(i in 1 :nrow(res1$dags)){
  res1_dags[[i]] <- t(matrix(res1$dags[i,], 7, 7, byrow = T))
}

# plot the DAG for each adj.matrix
par(mfrow=c(1,2))
for (i in 1:length(res1_dags)){
  qgraph(res1_dags[[i]], bidirectional=TRUE, color=c("white","salmon", "white", 
                                                 "white", "white", "white", "lightblue"), 
       layout = laymat, labels = varnames)
}

```


## 5. data. Choose two variables for which they should estimate the causal relationship
a. Cause variable: *ocdis*    
b. Outcome variable: *guilt*  


## 6. For the causal effect specified in 5
a. What is the true causal effect of the cause on the outcome variable based on your SCM?
path1  
path2
```{r}
 1 * 2 + 1 * 4 * 2 * 2 + 1 * 1 * 4
```

22?

b. Based on the true DAG, what linear regression model should be used to estimate the causal effect correctly?

```{r}
# mod_OCD <- lm(guilt ~ OCDdist + OCDint, data = OCDdata)
# summary(mod_OCD)
```
?

c. Estimate the causal effect with this regression model based on your generated data. To what extent is the true effect recovered?
```{r}
# cpdag_mat_OCD <- as(pc_OCD,"matrix")
# 
# # Each row is a DAG adjacency matrix in vector form (by rows)
# res_OCD <- pdag2allDags(cpdag_mat_OCD)
# 
# # We can get the adjacency matrix of an individual DAG using
# resOCD_dags <- list()
# for(i in 1:nrow(res_OCD$dags)){
#   resOCD_dags[[i]] <- t(matrix(res_OCD$dags[i,],7,7,byrow = TRUE))
# }
# 
# 
# ida(2,7, cov(OCDdata), pc_OCD@graph, verbose = TRUE)
```
The third CP-DAG does a good job?!
```{r}
# res_Adj <- resOCD_dags[[3]]
# 
# qgraph(res_Adj)
```



## 7. We will need a dichotomous cause variable
a. Make a dichotomized version of the cause variable from question 5&6 in your dataset, for example by assigning scores lower than the mean a 0 and the rest a 1.

```{r}
#Dichtomizing the effect variable OCDdist
# mean_OCDdist <- mean(OCDdistress)
# OCDdist_d <- rep(1, length(OCDdistress))
# OCDdist_d[which(OCDdistress < mean_OCDdist)] <- 0
#
# OCDdata <- cbind(OCDdata, OCDdist_d)

```

b. Use the correct model for the causal effect (the one from 6b) to estimate the causal effect again, but now with your dichotomized cause variable. Discuss the results.
```{r}
# mod_OCD_d <- lm(guilt ~ OCDdist_d + OCDint, data = OCDdata)
# summary(mod_OCD_d)
```

?


## 8. Prepare the data for the second causal inference assignment
a. Create a data frame with all your variables
```{r}
# save(OCDdata, file = "LoscherPark_Assign1_Data.RData")
```

b. Make a brief .txt file
See LoscherPark_Assign1.txt.



# Reference
  McNally, R.J., Mair, P., Mugno, B.L., & Riemann, B.C. (2017). Co-Morbid Obsessive–Compulsive Disorder and Depression: A Bayesian Network Approach. *Psychological Medicine*. *47*(7): 1204–14. https://doi.org/10.1017/S0033291716003287.

  Millet, B., Kochman, F., Gallarda, T., Krebs, M.O., Demonfaucon, F., Barrot, I., Bourdel, M.C., Olie, J.P., Loo, H., & Hantouche, E.G. (2004). Phenomenological and Comorbid Features Associated in Obsessive–Compulsive Disorder: Influence of Age of Onset. *Journal of Affective Disorders*. *79*(1-3): 241–46.

  Zandberg, L.J., Zang, Y., McLean, C.P., Yeh, R., Simpson, H.B., & Foa, E.B. (2015). Change in Obsessive-Compulsive Symptoms Mediates Subsequent Change in Depressive Symptoms During Exposure and Response Prevention. *Behaviour Research and Therapy*. *68*:76–81.






